---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r extract,cache=TRUE,results=FALSE}

unzip(zipfile = "repdata_data_activity.zip")
df1<-read.table(file = "activity.csv",sep = ",",header = TRUE)

```

Once obtained the data frame we check the head

```{r headCheck}
print(head(df))

```

Afterwards we can begin to preprocess the data by extracting the missing values and keep that data frame appart

```{r chaoMissing}

df<- df1[complete.cases(df1),]

```

Finally we can proceed with the calculation of the assignment requests.

## What is mean total number of steps taken per day?


Here is the code for reproducing that information

```{r histoSteps, cache=TRUE,results='asis'}

histogramFrame <- tapply(df$steps,df$date,sum)
histogramFrame <- data.frame(date = row.names(histogramFrame),steps.sum=histogramFrame)
row.names(histogramFrame)<-c()

barplot(histogramFrame$steps.sum ~ as.Date(histogramFrame$date),xlab = "Date", main = "Histogram of total number of \nsteps taken each day",ylab = "Total of steps per day")

```
 

 Here we calculate the median and the mean of steps taken per day
 
 ```{r mediMean, chache =TRUE}
 
 meanTable<-tapply(df$steps,df$date,mean)
 meanTable<-data.frame(date=row.names(meanTable),steps.mean=meanTable)
 row.names(meanTable)<-c()

 medianTable<-tapply(df$steps,df$date,median)
 medianTable<-data.frame(date=row.names(medianTable),steps.median=medianTable)
 row.names(medianTable)<-c()

 
 print(head(meanTable))
 print(head(medianTable))
 
 ```
 

Here is the plot of the average number of steps taken

```{r seriesTimeAvSteps,cache=TRUE}

plot(as.Date(meanTable$date),meanTable$steps.mean,type = 'l',pch=19,xlab = 'Date',ylab = "Daily mean steps")

```

## What is the average daily activity pattern?



```{r seriesTime}

meandaily<-tapply(df$steps,df$interval,mean)
meandaily<-data.frame(interval=as.numeric(row.names(meandaily)),steps.mean=meandaily)

plot(meandaily$interval,meandaily$steps.mean,xlab="Time in intervals of 5 minutes",ylab = "Mean of steps",pch=19,type = 'l')


```

It is possible to see that between the intervals 800 and 900 the mean of steps was superior than the other intervals.

## Imputing missing values



```{r missingVal,cache=TRUE,results=FALSE}
missingVals<-sum(1*!complete.cases(df1))

```

The amount of rows with no NA values is `r missingVals`. The strategy to replace the missing values consist on using the mean for each interval to replace the values in the column steps

```{r replacingNA,cache=TRUE}
library(plyr)

dfnew<-df1

dfnew$steps[is.na(df1$steps)]<-mapvalues(df1$interval[is.na(df1$steps)],meandaily$interval,meandaily$steps.mean)

```

### Histogram of the total number of steps taken each day after missing values are imputed


Now we make a new histogram based on the new data frame.

```{r histogramStrategy}

histoStrategy<-tapply(dfnew$steps, dfnew$date, sum)
histoStrategy <- data.frame(date=as.Date(row.names(histoStrategy)),steps.sum=histoStrategy)
row.names(histoStrategy)<-c()

barplot(steps.sum ~ as.Date(date),data=histoStrategy,xlab = "Date")

```
 
 As expected, the dates with non missing values conserve their height but the dates with plenty of missing values give heights to each column, making a huge difference due to the used strategy. 
 
 

## Are there differences in activity patterns between weekdays and weekends?


For the new plot, first we create a column with the weekdays and extract the weekend characteristic of the day.
 
 
 ```{r weekdaysPlot}
 library(lattice)
 
 dfnew$weekday<-weekdays(as.Date(dfnew$date))
 dfnew$weekday<-factor(dfnew$weekday,labels = c("weekday","weekday","weekday","weekday","weekday","weekend","weekend"),levels=unique(dfnew$weekday))
 
plotableDf<-aggregate(steps~interval+weekday,data = dfnew,mean)
xyplot(steps~interval|weekday,data = plotableDf,type="l",layout=c(1,2))
 
 ```

As presented, on weekend the activity is more spread and, in the mean, appears to have more activity. On other hand, the weekday presents the highest mean for steps.